---
title: "zero length groups"
author: "Romain François"
date: "12/04/2018"
output: md_document
---

```{r setup, include=FALSE}
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```

## install 

This is based on the `feature-341-zero-length-groups` branch: 

```r
devtools::install_github("tidyverse/dplyr", ref = "feature-341-zero-length-groups")
```

## Motivation

See the [full discussion](https://github.com/tidyverse/dplyr/issues/341) and the 
original [stackoverflow question](http://stackoverflow.com/questions/22523131). 

`dplyr` had the `drop` attribute in grouped data frames all along, but it did not really 
do anything about it, so in effect it has always been like if `drop` was set to TRUE. 

`drop=FALSE` is about keeping groups even if they have no data. 

```{r}
df <- data.frame( x = 1:2, g = factor(c("a", "b"), levels = c("a", "b", "c")))
df

tally(group_by(df, g, drop = TRUE))
tally(group_by(df, g, drop = FALSE))
```

With `drop=FALSE` the level "c" appears in the summary, with n=0. 

## Which groups to keep

The problem is that we don't always group on factors, and so the question that has kept the 
341 issue alive for 4 years is which groups. 

```{r}
df <- data.frame( 
  x  = 1:8,
  y  = rep(1:4, each=2),
  f1 = factor( rep(c("a", "b"), each = 4), levels = c("a", "b", "c")) 
)
df
```

- f1 has two distinct values in the data, but 3 levels
- y has 4 distinct values, but only two for the level "a" of f1 and 2 for the "b" level, and obviously 0 for the "c" level

```{r}
group_by(df, f1, drop = FALSE) %>% 
  summarise( n_distinct(y) )
```

So I'm not sure about what happens when we group by f1 and y, and when we group by y and f1. 

What we have currently in the `feature-341-zero-length-groups` is the cartesian product of 
levels of factors and unique values of non factors, so in that case we end with 12 groups
in both case

```{r}
group_by(df, y, f1, drop = FALSE) %>% tally() %>% arrange(desc(n))
group_by(df, f1, y, drop = FALSE) %>% tally() %>% arrange(desc(n))
```

In the thread issue, there are a lot of "we should keep all the groups only for factors", 
but I have no idea what to make of this. 



